The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks

### 2024-11-24 paper.Rmd
Added second simulation

```{r}
# Load required libraries
if (!require("reshape2")) install.packages("reshape2")
if (!require("ggplot2")) install.packages("ggplot2")
library(reshape2)
library(ggplot2)

@@

```{r}
# Reshape data for plotting
simulation_long <- melt(simulation, id.vars = "Time")

@@

```
```{r}
# Display the first few rows of the simulation data
knitr::kable(head(simulation), format = "html", caption = "Simulation Results (First Few Rows)")

@@

>>>>

```{r simulation_1, echo=FALSE}
# Load required libraries
library(reshape2)
library(ggplot2)

@@

```{r graph_1, echo=FALSE}
# Reshape data for plotting
simulation_long <- melt(simulation, id.vars = "Time")

@@

@@

## Radioactive Decay Simulation

This simulation models the stochastic decay of a single species (\( R \)), which represents a radioactive substance such as radon, radium, or plutonium. The decay process is governed by the following reaction:

\[
R \xrightarrow{c} 0
\]

where \( c \) is the decay rate constant that defines the likelihood of decay per unit time.

The Gillespie algorithm is used to simulate this system, capturing the inherent randomness of radioactive decay events. Unlike deterministic approaches, the Gillespie algorithm tracks individual decay events over time, providing an accurate representation of the stochastic nature of the process.

The goal of this simulation is to observe the time evolution of the remaining radioactive molecules (\( R \)) and visualize their decay over time.
```{r simulation_2, echo=FALSE}
# Define parameters
c <- 0.01  # Decay rate constant
state <- c(R = 1000)  # Initial radioactive substance
max_time <- 100  # Maximum simulation time

# Define propensity function
propensities <- function(state) {
  c * state["R"]
}

# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  time <- 0
  output <- data.frame(Time = numeric(0), R = numeric(0))
  output <- rbind(output, data.frame(Time = time, R = state["R"]))

  while (time < max_time) {
    # Calculate propensity
    prop <- propensities(state)
    if (prop == 0) break  # Stop if no reactions can occur

    # Time to next reaction
    delta_t <- rexp(1, rate = prop)
    time <- time + delta_t

    # Update state (decay reaction)
    state["R"] <- state["R"] - 1

    # Record the state
    output <- rbind(output, data.frame(Time = time, R = state["R"]))
  }

  return(output)
}

# Run the simulation
simulation <- simulate_gillespie(state, max_time)

```{r graph_2, echo=FALSE}
library(ggplot2)

# Plot results
ggplot(simulation, aes(x = Time, y = R)) +
  geom_line(color = "blue") +
  labs(
    title = "Radioactive Decay Simulation",
    x = "Time",
    y = "Remaining Radioactive Substance (R)"
  ) +
  theme_minimal()

### Observations

1. **Exponential Decay**:
   - The radioactive substance (\( R \)) decays exponentially over time, as expected from the properties of radioactive decay. The number of decay events per unit time decreases as the quantity of \( R \) diminishes.
   - This behavior is reflected in the propensity function \( a(R) = c \cdot R \), where the decay rate is proportional to the remaining amount of \( R \).

2. **Stochastic Fluctuations**:
   - Unlike a smooth deterministic decay curve, the simulation shows small fluctuations in the timing of decay events. These fluctuations arise from the stochastic nature of the Gillespie algorithm, which models each decay event as a probabilistic process.

3. **Time Steps and Decay Events**:
   - The time intervals between decay events (\( \Delta t \)) grow longer as the number of \( R \) molecules decreases. This is because the propensity (\( a(R) \)) becomes smaller, leading to larger expected time intervals between events.

### Biological/Physical Significance

- **Radioactive Half-Life**:
  - The exponential decay observed in this simulation is consistent with the concept of half-life in radioactive decay. While this model does not explicitly compute the half-life, it demonstrates how the decay rate slows down as the substance depletes.

- **Real-World Applications**:
  - Radioactive decay processes are critical in many fields, including nuclear physics, medicine (e.g., radiotherapy), and environmental science (e.g., radon gas monitoring). Understanding stochastic decay helps refine predictions in low-count systems where randomness plays a significant role.
